 

### 2021-07-07 vertical-thermal-review.Rmd
added citations, removed a few words

**Forest responses and feedbacks to climate change will have a critical influence on the future of Earth's climate.**
Globally, temperatures have risen # $1.2^\circ$C since the 1850-1900 period [@rohdeBerkeleyEarthLand2020] and are expected to rise an additional #-#$^\circ$C by YEAR (IPCC 2021), accompanied by increasing severity and frequency of heat waves [@meehlMoreIntenseMore2004] and hotter droughts [@trenberthGlobalWarmingChanges2014].
These changes are expected to have profound effects on tree metabolism and forest ecosystem function (@breshearsUnderappreciatedPlantVulnerabilities; IPBES report), including altered rates of photosynthesis and respiration [@breshearsUnderappreciatedPlantVulnerabilities 2021; @corlettImpactsWarmingTropical2011], foliar damage during heat waves [@osullivanThermalLimitsLeaf2017; @corlettImpactsWarmingTropical2011], and reduced growth and elevated mortality during drought [@teskeyResponsesTreeSpecies2015; @breshearsUnderappreciatedPlantVulnerabilities]. Collectively, these impacts may lead to reduced forest C sequestration, as may already be occurring in some forests [@oishiWarmerTemperaturesReduce2018; @sullivanLongtermThermalSensitivity2020].The resulting feedbacks to the climate system, decreased C storage, and changes in albedo and hydrology will in turn have a critical influence on the future trajectory of climate change [@bonan_ecological_2016, Bonan 2008], yet this influence remains extremely uncertain [REFS].

**Rising temperatures are likely to have differential impacts on trees of different size and canopy position, in large part because small understory trees exist in a microenvironment that is substantially buffered by the more exposed canopy trees [@zellweger_seasonal_2019; @davisMicroclimaticBufferingForests2019].**
Forests are vertically and horizontally stratified, with trees of different sizes through various successional stages. 
Overstory trees form canopies that play a crucial role in moderating forest macroclimates [@ozanneBiodiversityMeetsAtmosphere2003; @nakamuraForestsTheirCanopies2017], as well as microclimates by buffering understories from extreme temperatures and other macroclimatic conditions [@zellweger_seasonal_2019].
This creates a vertical stratification of biophysical environmental conditions such as temperature, light, wind, humidity and CO~2~ that influences leaf traits, thermoregulation and metabolism along the gradient, with implications for whole plant performance [@fauset_differences_2018; @michaletzEnergeticCarbonEconomic2016].
Despite the fact that this vertical gradient inevitably shapes nearly every aspect of plant metabolism, demography, and ecology, we lack comprehensive understanding of these gradients [but see @niinemetsPhotosyntheticAcclimationSimultaneous2004].
Importantly, this limits our ability to understand how warming temperatures will affect leaf-level metabolism, whole-plant performance, and, in turn, forest ecology and biodiversity, energy balance, ecosystem function, and biosphere-atmosphere interactions.

@@

**Wind speeds are also higher at the top of the canopy, owing to the effect of buffering with canopy height [@juckerCanopyStructureTopography2018].**
Wind speed increase with height across forest types from closed-canopy forests [@barnardSeasonalVariationCanopy2016; @mcgregor_tree_nodate; @juckerCanopyStructureTopography2018; @mullerEvidenceEfficientNonevaporative2021] to open savannas [@curtis_intracanopy_2019], including all 22 forested NEON sites (Figs. `r fig_NEON_vertical`e, `r SIfig_NEON_all`). 
Specifically, across the NEON sites, July maximum daily wind speeds averaged up to 4.1 m s^-1^ higher at the top of the NEON vertical profile than at the bottom, where maximum wind speeds averaged $\le$ 1.1 m s^-1^ (Figs. `r fig_NEON_vertical`, `r SIfig_NEON_all`).
The implication is that upper canopy leaves have substantially higher boundary layer conductance ($g_a$) and associated potential for loss of both water and heat (see next section).

@@

**Air temperature, $T_{air}$, often shows little variation across the vertical gradient, but under certain circumstances it can be significantly buffered by forest canopies (Figs. `r fig_NEON_vertical`, `r SIfig_NEON_all`).**
The $T_{air}$ gradient along canopies and understories varies with factors such as radiation fluxes, local topography, vertical air mixing, cloud cover and vegetation [*Defrenne et al. 2021*]. 
Typically, dense canopies buffer understories from maximum $T_{air}$ relatively more than open canopies -- *i.e.*, dense canopy understories can have cooler maximum daytime $T_{air}$ and warmer minimum $T_{air}$ during the night than open canopy understories or nearby clearings [*Defrenne et al. 2021*; @vonarxSpatiotemporalEffectsForest2012a; @defrenneGlobalBufferingTemperatures2019; @davisMicroclimaticBufferingForests2019; @zellweger_seasonal_2019; @ramboCanopyMicroclimateResponse2009]. 
Notably, this pattern can be reversed in open canopy forests or sparse stands, where below-canopy maximum $T_{air}$ can be warmer than canopy $T_{air}$ due to turbulent air mixing and thermal radiative flux from the soil and the canopy that is intercepted by lower-canopy layers, analogous to 'canopy greenhouse effect' [e.g., OSBS, Figs. `r fig_NEON_vertical`g, `r SIfig_NEON_all`b, @banerjeeExplainingConvectorEffect2017; @curtis_intracanopy_2019;@zweifelMiddayStomatalClosure2002; @hadleyInfluenceKrummholzMat1987; @ramboCanopyMicroclimateResponse2009; @hardwickRelationshipLeafArea2015].

@@

Leaf temperature ($T_{leaf}$) is strongly tied to $T_{air}$, and is further influenced by the other biophysical drivers reviewed above and by leaf traits (see next section), with leaves adjusted to their environment so as to maintain, when possible, an optimal $T_{leaf}$ for carbon assimilation and metabolic processes [@michaletzPlantThermoregulationEnergetics2015; @perezPhotosyntheticHeatTolerances2020]. 
Leaves can be cooler than air at high $T_{air}$ and warmer than air at low $T_{air}$ (limited homeothermy), and may sometimes be drastically warmer than $T_{air}$ (megathermy).
Often, however, $T_{leaf}$ tracks $T_{air}$ [poikilothermy, @cavaleriColdbloodedForestsWarming2020; @drakeNoEvidenceHomeostatic2020].

@@

$Ga$ and $Gs$ are dominant factors in $T_{leaf}$ flux. $Ga$ increases with wind speed which also in turn increases $gs$, thereby reduces $T_{leaf}$-$T_{air}$ (Fig. `r fig_leaf_T`b). Therefore, high windspeeds in upper canopies such as in NEON sites with windspeeds ~4 ms (Fig. `r fig_NEON_vertical`) are capable of lowering $T_{leaf}-T_{air}$ that are exposed to high radiation loads, upto ~ 4$^\circ$C (Fig. `r fig_leaf_T`b)[@bonan_ecological_2016; @PhotosyntheticAcclimationSimultaneous2004]. Low wind speeds typically decouple leaves from $T_{air}$, more so when solar radiation is high and $gs$ is low. Therefore, along with windspeed $T_{leaf}$ can fluctuate throughout the day. Additionally, $T_{leaf}-T_{air}$ increases with humidity (Fig. `r fig_leaf_T`c) because reduced $ga$ stifles heat dissapation through $E$ and $gs$.

@@

**Within forests, vertical gradients in the propensity for $T_{leaf}$ to be elevated above $T_{air}$ are expected to vary with micrometeorlogical conditions (Fig. `r fig_leaf_T`).**
$T_{leaf}$ generally tracks $T_{air}$, where vertical gradient of $T_{air}$ depends on canopy structure and $LAI$ [@leuzingerTreeSpeciesDiversity2007;@mullerEvidenceEfficientNonevaporative2021]. Closed canopies, and high $LAI$ canopies act as an umbrella, absorbing most of the incoming radiation and preventing vertical air mixing in the understory microclimate. The intensity of incoming radiation and $VPD$ in the upper canopy can limit $gs$. Therefore, we would expect a stronger $T_{leaf}$ gradient where upper canopy leaves can be much warmer than below canopies layers [@millerOnlySunlitLeaves2021]. However, higher windspeeds and lower RH (Fig. `r fig_NEON_vertical`), along with adaptive leaf traits that reduce $g_a$ and increase $g_s$ (reviewed below), give them lower sensitivity to intercepted radiation (Fig. `r fig_leaf_T`a) and higher rates of heat loss (Fig. `r fig_leaf_T`b-e).
On the other hand, open canopies, or lower canopy $LAI$ allow more vertical air mixing and sunlight into below canopies, that can either neutralize $T_{leaf}$ gradient or in the case of 'canopy greenhouse effect' or even sunflecks, can elevate below canopy $T_{leaf}$-$T_{air}$ of shade leaves more than upper canopy due to the propensity of shade leaves as mentioned above [@hardwickRelationshipLeafArea2015; @schymanskiStomatalControlLeaf2013].

@@

**Leaf traits determine the capacity of leaves to thermoregulate [@michaletzEnergeticCarbonEconomic2016], and as such are critical to optamize leaf metabolism (see following section) across forest vertical strata [@michaletzPlantThermoregulationEnergetics2015; @zwienieckiHydraulicLimitationsImposed2004]**
Leaf morphology adjusts to the local microenvironment during leaf expansion [@zwienieckiHydraulicLimitationsImposed2004], and the ratio of direct to diffused light and associated water demands determines the differentiation of traits of sun and shade leaves [@casasSunShadeLeaves2011; @zwienieckiHydraulicLimitationsImposed2004; @keenanGlobalLeafTrait2016; @niinemetsLeafAgeDependent2016a]. 
This dichotomy in leaves is observed along the vertical gradient where canopy-leaves vary dramatically from canopy-interior and understory due to the differences in biophysical conditions (Figs. `r fig_NEON_vertical`, `r SIfig_NEON_all`).

@@

Within-canopy, leaf traits vary along the canopy light gradient [*i.e.,* in sun and shade leaves, @casasSunShadeLeaves2011] at a leaf anatomical, structural and biochemical level [Table `r table_leaf_traits`, @sackHowStrongIntracanopy2006; @niinemetsWorldwideAnalysisWithincanopy2015; @chenLeafEconomicsSpectrum2020]. 
This variation shapes vertical $T_{leaf}$ patterns [@fauset_differences_2018; @michaletzEnergeticCarbonEconomic2016]. 
Further research to characterize trait relationships and responses through the vertical canopy is needed, because much of our understanding of trait coordination is based on the leaf economics spectrum concept (Wright et al. 2004) was developed using sun leaves [@chenLeafEconomicsSpectrum2020; @keenanGlobalLeafTrait2016].

@@

Sun leaves in the exposed canopy have traits that help avoid heat and water stress than shade leaves where, for instance, being smaller, hairier, with higher vein density [@leighInfluenceLeafSize2017; @vogelConvectiveCoolingLow1970; @sackHowStrongIntracanopy2006; @zwienieckiHydraulicLimitationsImposed200], and steeper leaf-angle [@niinemetsAdjustmentFoliageStructure1998, *REF*]. These traits [see Table 1] collectively aid heat dissapation and regulate $T_{leaf}$-photosynthetic balance. Whereas, shade leaves have traits that maximize light capture in the diffused-light environment with sunflecks, for example lower LMA, but are more likely to get overheated than sun leaves due to greater thermal capacitance [@mathurPhotosyntheticEfficiencySun2018a; @leighInfluenceLeafSize2017; @casasSunShadeLeaves2011; @schymanskiStomatalControlLeaf2013a]. In open canopies where light is comparitively homogenous, exposed leaves are primarily shaped by $T_{air_{max}}$ and $VPD$ stress, rather than solely by light [@mediavillaFoliarPlasticityRelated2019a].

@@

Biochemical protection against foliage light and heat damage also increases with irradiance, and thus tends to be higher in the upper canopy than in the understory. 
Caratenoids and xanthophyll cycle pigments--violaxanthin, antheraxanthin and zeaxanthin [$VAZ$]--can play a role in antioxidant scavenging and converting excess excitation energy into heat to reduce photoinhibition [@niinemetsPhotosynthesisResourceDistribution2007a; @mathurPhotosyntheticEfficiencySun2018a; @niinemetsAcclimationHighIrradiance1998]. This is also important for leaf acclimation to higher temperature if high temperature impairs the photochemical energy dissipation pathway [@TemperaturedependentAdjustmentThermal1996].
Elemental traits, such as $N_{area}$ and $P_{area}$, contribute to high activities of leaf metabolism, including respiration [@meirLeafRespirationTwo2001;@weerasingheCanopyPositionAffects2014] and photosynthetic processes in sun leaves, where $V_{cmax}$ and $J_{max}$ [see next section]-- are linearly related to within-canopy foliar $N_{area}$ and light gradient (Table 1)[@scartazzaInvestigatingEuropeanBeech2016a;@niinemetsWithincanopyVariationRate2004;@weerasingheCanopyPositionAffects2014]. Evidence also shows that within canopy leaf nitrogen distribution is regulated to prevent photoinhibition in shade leaves during sunlecks, and direct light for sun leaves [@kitaoCanopyNitrogenDistribution2018]. Hence, although we would expect elemental traits to influence vertical foliar thermal sensitivity, this remains unclear as @zhuPlasticityPhotosyntheticHeat2018 and @osullivanThermalLimitsLeaf2017 observed no relation between foliar thermal tolerance and total foliar concentrations of N and P.  Further, higher heat tolerance in exposed canopies is facilitated by isoprenes in species capable of emitting isoprenes, providing these species with a selective advantage in a warming world [*REF*].

@@

Leaf age affects vulnerability to heat stress independent of light environment [e.g. Marias et al. 2017], and has a strong effect on metabolism [@kikuzawaSynthesisRelationshipsLeaf2006]. Sun leaves have shorter lifespans than shade leaves, and thus upper canopy leaves are on average younger than subcanopy leaves, they therefore operate closer to their photosynthetic optimum . However, because leaf turnover rates of sun leaves in the upper canopy are faster than in the subcanopy,  damage does not really accumulate [@millerOnlySunlitLeaves2021]. The effect of leaf age on heat tolerance once leaves are fully expanded and mature, is XXX.

@@

Widely, across forest species, the pattern of fundamental trait adaptation to light gradients is similar to within-canopy traits [@rozendaalPlasticityLeafTraits2006; @thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012]. However, trait plasticity can vary several-folds between  multi-species canopy and in understory species,  based on their inherent specie-specific traits and responses to environmental conditions. Spatio-temporal gradients, offer another axis for trait variation, for instance in early- and late-successional species [Ishii and Asano, 2007; @ishiiExploringRelationshipsCanopy2004a]. Therefore,  $T_{leaf}$ differs at an intra- and inter-specific levels with traits, where species with smaller leaves and larger $gs$ will be less vulnerable to temperature increase than species with larger leaves [@fauset_differences_2018; @songTreeSurfaceTemperature2020].

@@

**Photosynthesis is generally higher in exposed canopy positions-- a fact that is both theoretically expected and observed in numerous field studies (Table 2; Niinemets 2007 ; Kenzo et al. 2015 ; Slot et al. 2019; Chen et al. 2020).**
This is primarily driven by the greater light available to sun-leaves. Photosynthesis increases with increasing irradiance, and saturates at quantum flux density of 400 $m^{-2}$ $s^{-1}$ to 700nm $m^{-2}$ $s^{-1}$.

In response to the greater light availability, sun-leaves have traits allowing greater photosynthetic rates at high light under sufficient nutrient and water conditions [Table 1; Table 2]. Light saturation levels differ between leaves and canopies with leaves saturating at lower light level than canopies; while a certain amount of light saturates the photosynthetic capacity of the leaf, increasing the amount of light in the canopy may increase photosynthesis in the shaded part of the canopy.The vertical gradient in photosynthetic rates is also influenced by available concentration of N and P, and water availability.

Temperature can affect photosynthesis via direct and indirect pathways--directly, by altering photosynthetic enzyme activity and electron transport chain, and indirectly through increase in $T_{leaf}$ from $gs$ closure that is driven by biophysical factors [@lloydEffectsRisingTemperatures2008]. Photosynthesis has a peaked response to temperature, commonly found to be maximized at the prevalent ambient growing season conditions [Tan et al. 2017; @slotSituTemperatureResponse2017]. Beyond the optimum, photosynthesis decreases as a result of stomatal closure and increasing $T_{leaf}$ (e.g. @slotSituTemperatureResponse2017; Smith et al. 2020 [temperate/boreal refs needed), and eventually due to biochemical constraints (@kumarathungeAcclimationAdaptationComponents2019a; e.g. Sage & Kubien 2007; Varhammer et al. 2015 ). Leaf heat stress at $40-60{^\circ}C$ can cause irreversible damage to photosystem [II], leading to leaf necrosis and eventually leaf death [Baker 2008; @feeley_thermal_2020].

**We have very little evidence as to how the temperature sensitivity of photosynthesis compares between sun and shade leaves.**
Based on consistent positive relationships between growth temperature and $T_{Opt}$ across sites and seasons (e.g. Tan et al. 2017; Kumarathunge et al. 2019), one might expect sun leaves to have a stronger temperature-dependence and higher temperature optima than shade leaves (Fig. 14.5) in Campbell and Norman 1998; *revisit this!*; @niinemetsShapeLeafPhotosynthetic1999; @niinemetsPhotosyntheticAcclimationSimultaneous2004]. However, such a trend is not apparent among recent studies, where for 3 species in Panama, the optimum temperature for sun leaves tended to be moderatly, but not significantly higher than that of[ @slotPhotosyntheticHeatTolerance2019a; @hernandezSimilarTemperatureDependence2020]. Similarly, $T_{opt}$ of $V_{max}$ and $J_{max}$ didnot differ systematically between tropical sun- and shade-leaves [@hernandezSimilarTemperatureDependence2020], and along vertical gradients of tropical and temperate trees [@mau_temperate_2018; @millerOnlySunlitLeaves2021;@mau_temperate_2018].

>>>>

**Forest responses and feedbacks to climate change will have a critical influence on the future of Earth's climate.**
Globally, temperatures have risen $1.2^\circ$C since the 1850-1900 period [@rohdeBerkeleyEarthLand2020] and are expected to rise over pre-industrial levels $1.6 - 4.3^{\circ}C$ by 2100 (IPCC 2014), accompanied by increasing severity and frequency of heat waves [@meehlMoreIntenseMore2004] and hotter droughts [@trenberthGlobalWarmingChanges2014b].
These changes are expected to have profound effects on tree metabolism and forest ecosystem function (@breshearsUnderappreciatedPlantVulnerabilities 2021; IPBES report), including altered rates of photosynthesis and respiration [@breshearsUnderappreciatedPlantVulnerabilities 2021; @corlettImpactsWarmingTropical2011], foliar damage during heat waves [@osullivanThermalLimitsLeaf2017; @corlettImpactsWarmingTropical2011], and reduced growth and elevated mortality during drought [@teskeyResponsesTreeSpecies2015; @breshearsUnderappreciatedPlantVulnerabilities]. Collectively, these impacts may lead to reduced forest C sequestration, as may already be occurring in some forests [@oishiWarmerTemperaturesReduce2018; @sullivanLongtermThermalSensitivity2020]. The resulting feedbacks to the climate system, decreased C storage, and changes in albedo and hydrology will in turn have a critical influence on the future trajectory of climate change [@bonan_ecological_2016, Bonan 2008], yet this influence remains extremely uncertain [REFS].

**Rising temperatures are likely to have differential impacts on trees of different size and canopy position, in large part because small understory trees exist in a microenvironment that is substantially buffered by the more exposed canopy trees [@zellweger_seasonal_2019; @davisMicroclimaticBufferingForests2019].**
Forests are vertically and horizontally stratified, with trees of different sizes through various successional stages. 
Overstory trees form canopies that play a crucial role in moderating forest macroclimates [@ozanneBiodiversityMeetsAtmosphere2003; @nakamuraForestsTheirCanopies2017], as well as microclimates by buffering understories from extreme temperatures and other macroclimatic conditions [@zellweger_seasonal_2019].
This creates a vertical stratification of biophysical environmental conditions such as temperature, light, wind, humidity and CO~2~ that influences leaf traits, thermoregulation and metabolism along the gradient, with implications for whole plant performance [@fauset_differences_2018; @michaletzEnergeticCarbonEconomic2016].
Despite the fact that this vertical gradient inevitably shapes nearly every aspect of plant metabolism, demography, and ecology, we lack comprehensive understanding of these gradients [but see @niinemetsPhotosyntheticAcclimationSimultaneous2004b].
Importantly, this limits our ability to understand how warming temperatures will affect leaf-level metabolism, whole-plant performance, and, in turn, forest ecology and biodiversity, energy balance, ecosystem function, and biosphere-atmosphere interactions.

@@

**Wind speeds are also higher at the top of the canopy, owing to the effect of buffering with canopy height [@juckerCanopyStructureTopography2018].**
Wind speed increase with height across forest types from closed-canopy forests [@barnardSeasonalVariationCanopy2016a; @mcgregor_tree_nodate; @juckerCanopyStructureTopography2018; @mullerEvidenceEfficientNonevaporative2021] to open savannas [@curtis_intracanopy_2019], including all 22 forested NEON sites (Figs. `r fig_NEON_vertical`e, `r SIfig_NEON_all`). 
Specifically, across the NEON sites, July maximum daily wind speeds averaged up to 4.1 m s^-1^ higher at the top of the NEON vertical profile than at the bottom, where maximum wind speeds averaged $\le$ 1.1 m s^-1^ (Figs. `r fig_NEON_vertical`, `r SIfig_NEON_all`).
The implication is that upper canopy leaves have substantially higher boundary layer conductance ($g_a$) and associated potential for loss of both water and heat (see next section).

@@

**Air temperature, $T_{air}$, often shows little variation across the vertical gradient, but under certain circumstances it can be significantly buffered by forest canopies (Figs. `r fig_NEON_vertical`, `r SIfig_NEON_all`).**
The $T_{air}$ gradient along canopies and understories varies with factors such as radiation fluxes, local topography, vertical air mixing, cloud cover and vegetation [@frenneForestMicroclimatesClimate2021]. 
Typically, dense canopies buffer understories from maximum $T_{air}$ relatively more than open canopies -- *i.e.*, dense canopy understories can have cooler maximum daytime $T_{air}$ and warmer minimum $T_{air}$ during the night than open canopy understories or nearby clearings [@frenneForestMicroclimatesClimate2021; @vonarxSpatiotemporalEffectsForest2012a; @defrenneGlobalBufferingTemperatures2019; @davisMicroclimaticBufferingForests2019; @zellweger_seasonal_2019; @ramboCanopyMicroclimateResponse2009]. 
Notably, this pattern can be reversed in open canopy forests or sparse stands, where below-canopy maximum $T_{air}$ can be warmer than canopy $T_{air}$ due to turbulent air mixing and thermal radiative flux from the soil and the canopy that is intercepted by lower-canopy layers, analogous to 'canopy greenhouse effect' [e.g., OSBS, Figs. `r fig_NEON_vertical`g, `r SIfig_NEON_all`b, @banerjeeExplainingConvectorEffect2017; @curtis_intracanopy_2019;@zweifelMiddayStomatalClosure2002; @hadleyInfluenceKrummholzMat1987; @ramboCanopyMicroclimateResponse2009; @hardwickRelationshipLeafArea2015].

@@

Leaf temperature ($T_{leaf}$) is strongly tied to $T_{air}$, and is further influenced by the other biophysical drivers reviewed above and by leaf traits (see next section), with leaves adjusted to their environment so as to maintain, when possible, an optimal $T_{leaf}$ for carbon assimilation and metabolic processes [@michaletzPlantThermoregulationEnergetics2015; @perezPhotosyntheticHeatTolerances2020]. 
Leaves can often be either cooler than air at high $T_{air}$ or warmer than air at low $T_{air}$ (limited homeothermy), and may sometimes be drastically warmer than $T_{air}$ (megathermy) [@doughtyAreTropicalForests2008].
However, rarely is $T_{leaf}$ equal to $T_{air}$ [poikilothermy, @cavaleriColdbloodedForestsWarming2020; @drakeNoEvidenceHomeostatic2020].

@@

$Ga$ and $Gs$ are dominant factors in $T_{leaf}$ flux. $Ga$ increases with wind speed which also in turn increases $gs$, thereby reduces $T_{leaf}$-$T_{air}$ (Fig. `r fig_leaf_T`b). Therefore, high windspeeds in upper canopies such as in NEON sites with windspeeds ~4 ms (Fig. `r fig_NEON_vertical`) are capable of lowering $T_{leaf}-T_{air}$ that are exposed to high radiation loads, upto ~ 4$^\circ$C (Fig. `r fig_leaf_T`b)[@bonan_ecological_2016; @PhotosyntheticAcclimationSimultaneous2004]. Low wind speeds typically decouple leaves from $T_{air}$, more so when solar radiation is high and $gs$ is low. Additionally, $T_{leaf}-T_{air}$ increases with humidity (Fig. `r fig_leaf_T`c) because reduced $ga$ stifles $E$ and $gs$ that dissapate heat [@songTreeSurfaceTemperature2020].

@@

**Within forests, vertical gradients in the propensity for $T_{leaf}$ to be elevated above $T_{air}$ are expected to vary with micrometeorlogical conditions (Fig. `r fig_leaf_T`).**
$T_{leaf}$ generally tracks $T_{air}$, where vertical gradient of $T_{air}$ depends on canopy structure and $LAI$ [@leuzingerTreeSpeciesDiversity2007;@mullerEvidenceEfficientNonevaporative2021]. Closed canopies, and high $LAI$ canopies act as an umbrella, absorbing most of the incoming radiation and preventing vertical air mixing in sub canopies and the understory microclimate. The intensity of incoming radiation and $VPD$ in the upper canopy can limit $gs$, therefore, we would expect a stronger $T_{leaf}$ gradient in closed canopies where upper canopy leaves can be much warmer than below canopies layers [@millerOnlySunlitLeaves2021]. However, higher windspeeds and lower RH (Fig. `r fig_NEON_vertical`), along with adaptive leaf traits that reduce $g_a$ and increase $g_s$ (reviewed below), give them lower sensitivity to intercepted radiation (Fig. `r fig_leaf_T`a) and higher rates of heat loss (Fig. `r fig_leaf_T`b-e) [@millerOnlySunlitLeaves2021].
On the other hand, open canopies, or lower canopy $LAI$ allow more turbulent air mixing and sunlight into below canopies, that can either neutralize $T_{leaf}$ gradient or in the case of 'canopy greenhouse effect' or even sunflecks, can elevate below canopy $T_{leaf}$-$T_{air}$ of shade leaves more than upper canopy due to the propensity of shade leaves [@hardwickRelationshipLeafArea2015; @schymanskiStomatalControlLeaf2013].

@@

**Leaf traits determine the capacity of leaves to thermoregulate [@michaletzEnergeticCarbonEconomic2016], and as such are critical to optamize leaf metabolism (see following section) across forest vertical strata [@michaletzPlantThermoregulationEnergetics2015; @zwienieckiHydraulicLimitationsImposed2004]**
Leaf morphology adjusts to the local microenvironment during leaf expansion [@zwienieckiHydraulicLimitationsImposed2004], and the ratio of direct to diffused light and associated water demands determines the differentiation of traits of sun and shade leaves [@casasSunShadeLeaves2011; @zwienieckiHydraulicLimitationsImposed2004; @keenanGlobalLeafTrait2016]. 
This dichotomy in leaves is observed along the vertical gradient where canopy-leaves vary dramatically from canopy-interior and understory due to the differences in biophysical conditions (Figs. `r fig_NEON_vertical`, `r SIfig_NEON_all`).

@@

Within-canopy, leaf traits vary along the canopy light gradient [*i.e.,* in sun and shade leaves, @casasSunShadeLeaves2011] at a leaf anatomical, structural and biochemical level [Table `r table_leaf_traits`, @sackHowStrongIntracanopy2006; @niinemetsWorldwideAnalysisWithincanopy2015; @chenLeafEconomicsSpectrum2020]. 
This variation shapes vertical $T_{leaf}$ patterns [Table 1; @fauset_differences_2018; @michaletzEnergeticCarbonEconomic2016]. 
Further, research to characterize trait relationships and responses through the vertical canopy is needed, because much of our understanding of trait coordination is based on the leaf economics spectrum concept (Wright et al. 2004) was developed using sun leaves [@chenLeafEconomicsSpectrum2020; @keenanGlobalLeafTrait2016].

@@

Sun leaves in the exposed canopy have traits that help avoid heat and water stress than shade leaves, for instance, by being smaller, hairier, with higher vein density [@leighInfluenceLeafSize2017; @vogelConvectiveCoolingLow1970; @sackHowStrongIntracanopy2006; @zwienieckiHydraulicLimitationsImposed200], and steeper leaf-angle [@niinemetsAdjustmentFoliageStructure1998, *REF*]. These traits [see Table 1] collectively aid heat dissapation and regulate $T_{leaf}$-photosynthetic balance. Whereas, shade leaves have traits that maximize light capture in the diffused-light environment with sunflecks, for example lower LMA, but are more likely to get overheated than sun leaves due to greater thermal capacitance [@mathurPhotosyntheticEfficiencySun2018a; @leighInfluenceLeafSize2017; @casasSunShadeLeaves2011; @schymanskiStomatalControlLeaf2013b]. In open canopies where light is comparitively homogenous, exposed leaves are primarily shaped by $T_{air_{max}}$ and $VPD$ stress, rather than solely by light [@mediavillaFoliarPlasticityRelated2019a].

@@

Biochemical protection against foliage light and heat damage also increases with irradiance, and thus tends to be higher in the upper canopy than in the understory. 
Caratenoids and xanthophyll cycle pigments--violaxanthin, antheraxanthin and zeaxanthin [$VAZ$]--can play a role in antioxidant scavenging and converting excess excitation energy into heat to reduce photoinhibition [@niinemetsPhotosynthesisResourceDistribution2007a; @mathurPhotosyntheticEfficiencySun2018a; @niinemetsAcclimationHighIrradiance1998]. This is also important for leaf acclimation to higher temperature if high temperature impairs the photochemical energy dissipation pathway [@TemperaturedependentAdjustmentThermal1996].

Elemental traits, such as $N_{area}$ and $P_{area}$, contribute to activities of high leaf metabolism, including respiration [@meirLeafRespirationTwo2001;@weerasingheCanopyPositionAffects2014] and photosynthetic processes in sun leaves [Table 1; @scartazzaInvestigatingEuropeanBeech2016a;@niinemetsWithincanopyVariationRate2004;@weerasingheCanopyPositionAffects2014]. Evidence also shows that within canopy leaf nitrogen distribution is regulated to prevent photoinhibition in shade leaves during sunlecks, and direct light for sun leaves [@kitaoCanopyNitrogenDistribution2018]. Hence, although we would expect elemental traits to influence vertical foliar thermal sensitivity, this remains unclear as @zhuPlasticityPhotosyntheticHeat2018 and @osullivanThermalLimitsLeaf2017 observed no relation between foliar thermal tolerance and total foliar concentrations of N and P.  Further, higher heat tolerance in exposed canopies is facilitated by isoprenes in species capable of emitting isoprenes, providing these species with a selective advantage in a warming world [*REF*].

@@

Leaf age affects vulnerability to heat stress independent of light environment [e.g. Marias et al. 2017], and has a strong effect on metabolism [@kikuzawaSynthesisRelationshipsLeaf2006], including photosynthesis and respiration, and their temperature sensitivities [@zhouLeafageEffectsTemperature2015; @niinemetsLeafAgeDependent2016]. Sun leaves have shorter lifespans than shade leaves, and thus upper canopy leaves are on average younger than subcanopy leaves, therefore they operate closer to their photosynthetic optimum. However, because leaf turnover rates of sun leaves in the upper canopy are faster than in the subcanopy, damage does not really accumulate [@millerOnlySunlitLeaves2021]. The effect of leaf age on heat tolerance once leaves are fully expanded and mature, is that heat tolerance decreases with increasing leaf age [@zhouLeafageEffectsTemperature2015; REFS], along with photosynthesis and $gs$ [@zhouLeafageEffectsTemperature2015; @niinemetsLeafAgeDependent2016]. In evergreen species, temperature responses of photosynthesis and temperature sensitivity of respiration ($Q_{10}$) have also been observed to be greater in younger than older leaves [@zhouLeafageEffectsTemperature2015]. Additionally, because older leaves reach maximum $T_{leaf}$ faster than younger leaves and can occupy a large part of the canopy, this would affect vertical gradient of thermal sensitivity [@zhouLeafageEffectsTemperature2015].

@@

Generally, the pattern of fundamental trait adaptation to light gradients is across forest species is similar to within-canopy traits [@rozendaalPlasticityLeafTraits2006; @thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012]. However, the range in trait values can be several times greater in multi-species canopy than in understory species,  based on their inherent species-specific traits and responses to environmental conditions. Spatio-temporal gradients, offer another axis for trait variation, for instance in early- and late-successional species [Ishii and Asano, 2007; @ishiiExploringRelationshipsCanopy2004a]. Therefore,  patterns in $T_{leaf}$ are underpinned by both intra- and inter-specific variation in traits. For example, species with smaller leaves and larger $gs$ will be less vulnerable to temperature increase than species with larger leaves [@fauset_differences_2018; @songTreeSurfaceTemperature2020].

@@

**Photosynthetic capacity is generally higher in exposed canopy positions-- a fact that is both theoretically expected and observed in numerous field studies (Table 2; Niinemets 2007 ; Kenzo et al. 2015 ; Slot et al. 2019; Chen et al. 2020).**
This is primarily driven by the greater light available to sun-leaves that have traits [Table 1; Table 2] allowing greater photosynthetic rates at high light under sufficient nutrient and water conditions. Photosynthesis increases with irradiance, and saturates at quantum flux density of 400 $m^{-2}$ $s^{-1}$ to 700nm $m^{-2}$ $s^{-1}$. Light saturation levels differ between leaves and whole canopies with leaves saturating at lower light level than canopies; while a certain amount of light saturates the photosynthetic capacity of the leaf, increasing the amount of light in the canopy may increase photosynthesis in the shaded part of the canopy.

Temperature can affect photosynthesis via direct and indirect pathways--directly, by altering photosynthetic enzyme activity and electron transport chain, and indirectly through increased $VPD$ causing stomatal closure [@lloydEffectsRisingTemperatures2008]. Photosynthesis has a peaked response to temperature, with the peak commonly found to correspond to the prevalent ambient growing season conditions [Tan et al. 2017; @slotSituTemperatureResponse2017]. Beyond the optimum, photosynthesis decreases as a result of stomatal closure (e.g. @slotSituTemperatureResponse2017; Smith et al. 2020 [temperate/boreal refs needed), and eventually due to biochemical constraints (@kumarathungeAcclimationAdaptationComponents2019a; Sage & Kubien 2007; Varhammer et al. 2015). Irresversible damage to photosystem II, leading to leaf necrosis and eventually leaf death occurs at $T_{leaf}$ of $40-60^{\circ}C$
[Baker 2008; @feeley_thermal_2020].

**We have very little evidence as to how the temperature sensitivity of photosynthesis compares between sun and shade leaves.**
Based on consistent positive relationships between growth temperature and $T_{opt}$ across sites and seasons (e.g. Tan et al. 2017; Kumarathunge et al. 2019), one might expect sun leaves to have a stronger temperature-dependence and higher temperature optima than shade leaves (Fig. 14.5) in Campbell and Norman 1998; *revisit this!*; @niinemetsShapeLeafPhotosynthetic1999; @niinemetsPhotosyntheticAcclimationSimultaneous2004]. However, such a trend is not apparent among recent studies, where for 3 species in Panama, the optimum temperature for sun leaves tended to be moderatly, but not significantly higher than that of shade leaves [@slotPhotosyntheticHeatTolerance2019a; @hernandezSimilarTemperatureDependence2020]. Similarly, $T_{opt}$ of the RuBP carboxylation  ($V_{max}$) and regeneration rates ($J_{max}$) didnot differ systematically between tropical sun- and shade-leaves [@hernandezSimilarTemperatureDependence2020], and along vertical gradients of tropical and temperate trees [@mau_temperate_2018; @millerOnlySunlitLeaves2021;@mau_temperate_2018].
